---
title: "Robertson and Cao Project"
subtitle: "Teenage HPV Vaccination Coverage and Socioeconomic Factors"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---
# *Part 1*
## Data Set
  The data we chose to use is the National Immunization Survey (NIS) of 2022. It consists survey data that was collected to monitor vaccination coverage for teenagers age 13-17 in the United States of America. The surveys were conducted by the National Center for Immunization and Respiratory Diseases of the Center (NCIRD) for Disease Control and Prevention (CDC). The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.”
  The original data set is a fixed width file (FWF). In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and the each column representing the answer to a question. The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information. 
We have truncated the data set to include 26-selected columns for the ease of observation. We have also filtered the data set to include one year, 2022, which is the most recent year of data that is published. The provided survey data will require cleaning, as there are signs of human errors present. 
## Opening the Data Set

First, we will load the readr package to open the file
```{r}
library(readr) #Loading the package to read the fwf
```


Next, we will define the column widths (beginning and end positions) and applying the column names to each position.
```{r}
column_position<- fwf_positions(
  start = c(1, 89, 287, 295, 298, 314, 330, 331, 334, 336, 343, 344, 345,
            347, 355, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  end = c(5, 92, 288, 295, 313, 329, 330, 332, 334, 336, 343, 344, 346,
          347, 356, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR", "INCPORAR_I", 
                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
                "INS_BREAK_I")
)

```


Create the data frame "data" with read_fwf function. 
```{r}
data_frame <- read_fwf( "../products/NISTEENPUF22.DAT", 
                  col_positions = column_position, 
                  col_types = cols(.default = "c")
)
```

```{r}
#Read the statistics in the data file.
  summary(data_frame)
  str(data_frame)
```

The data will then be cleaned. As of now, the data set requires a library to translate the code.

The data will then be converted the data frame to a readable and downloadable .csv file for further analysis
```{r}
output_file<-"NIS_Teen_Data_2022.csv"
write.csv(data_frame, file = output_file, row.names = FALSE)
```





## Research Question
  We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S. Specifically, we are asking the question, What is the likelihood of a teenager being up-to-date for their HPV vaccinations in the U.S., based on socioeconomic factors and geographic distribution. The outcome of this study will be measured using the columns that pertain to HPV vaccination completion. These variables are provider-collected, and they state whether or not the teen is (1) up-to-date with 1+ HPV shot, (2) up-to-date with 2+ HPV shots, or (3) up-to-date with 3+ HPV shots (excluding all vaccinations post-survey). Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime. This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis.
  We aim to measure determinants of socioeconomic status that are engrained into the survey questions. These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status. Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12. Geographic distribution of the teens will be assessed by true state of residence. 
  We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location. This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.

## Proposed Analysis
	For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors. When performing this, we may choose to stratify by age or gender, if we find these to be confounders. The correlation analysis cannot be performed using a multivariate model because there may be collinearity present between many of the determinants for socioeconomic status. This will lead to spurious results. Instead, the correlation between vaccination status and each factor is examined separately. 
	Next, we may stratify this by state. Comparing these correlation values between states and regions may lead us to find patterns of healthcare accessibility in the U.S. It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.(1), but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 (2). More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic (3). Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

{{< pagebreak >}}

## References
1. Sciences, N. A. of, Engineering, & Medicine, and. (2018, March 1). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/ 
2. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M. (2022). Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study. SSM - population health, 18, 101082. https://doi.org/10.1016/j.ssmph.2022.101082
3. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024). Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates. BMJ Medicine, 3, 000726. doi: 10.1136/bmjmed-2023-000726